Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves

### 2025-07-07 ms/ms.tex
updating select model 2

We tested these hypotheses by comparing \aax{} among amphistomatous wild
tomato species (\emph{51}) from different native light habitats, grown
under simulated sun and shade light treatments, and measured under
contrasting light intensity (). We measured AA on 572 individual plants
from 29 populations (average of 9.86 replicates per light treatment)
using a recently developed method (\emph{46}). With this method, we
directly compare the photosynthetic rate of an untreated amphistomatous
leaf to that of the same leaf with gas exchange blocked through the
adaxial (upper) surface by transparent plastic, which we refer to as
`pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at
identical whole-leaf \(g_\text{sc}\), we measure \(A\) over a range of
\(g_\text{sc}\), inducing stomatal opening and closure by modulating
humidity (see Materials and Methods for further details). We estimated
`amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but with
modifications previously described in (\emph{46}) and here (Materials
and Methods). The native light intensity was represented by plant area
index (PAI \(\unit{\meter\squared\per\meter\squared}\)), estimated using
a global gridded X-m2 resolution data set derived from the Global
Ecosystem Dynamics Investigation {[}GEDI; (\emph{52}){]} and
georeferenced accession collection information from the Tomato Genetics
Resource Center. The growth light intensities were \ppfdequals{761} (sun
treatment) and \ppfdqty{115} (shade treatment) while all other
environment conditions were nearly identical (see materials and methods
in supplementary materials). The high and low measurements intensities
were \ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150}
(87.0:13.0 red:blue), respectively.

Consistent with biophysical theory of CO\(_2\) diffusion within leaves,
\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). \aax{} varied
substantially between measurement light intensities, growth light
intensities, and among populations (point to model comparison evidence).
Measured under high light intensity, \aax{} was consistently greater for
sun plants. The average \aax{} among populations in the shade treatment
was 0.039 (range: 0.012--0.109; 21 of 29 populations significant);
however, the same populations grown at high light intensity showed a
mean \aax{} of 0.050 (range: 0.024--0.120; 25 of 29 populations
significant). Contrary to the predictions of the assimilatory
hypothesis, \aax{} was greater in all populations under low measurement
light intensity for both sun and shade grown plants. The effect of low
light on \aax{} was more pronounced in the sun-grown plants, where
\aax{} was significantly greater under low measurement light intensity
in 28 of 29 populations compared to to 11 populations for shade-grown
plants. The overall average \aax{} of shade and sun grown plants
measured under low light intensity was 0.066 (range: 0.019--0.143; 27 of
29 populations significant) and 0.098 (range: 0.051--0.175; 29 of 29
populations significant), respectively. There was a slight tendency for
populations from more closed habitats (greater PAI), but \aax{} varied
widely among populations from open habitats (low PAI) regardless of
growth and measurement light intensities (Fig.~\ref{fig-aa}B).

@@

@@

\toprule
Species & TGRC accession & Latitude & Longitude & Elevation (mas) & $\mathrm{PAI}~(\unit{\meter\squared\per\meter\squared})$\\
\midrule
\em{\cellcolor{gray!10}{S. arcanum}} & \cellcolor{gray!10}{LA2172} & \cellcolor{gray!10}{-6.008} & \cellcolor{gray!10}{-78.858} & \cellcolor{gray!10}{662} & \cellcolor{gray!10}{0.70}\\
\em{S. cheesmaniae} & LA0429 & -0.644 & -90.329 & 800 & 0.60\\
\em{\cellcolor{gray!10}{S. cheesmaniae}} & \cellcolor{gray!10}{LA3124} & \cellcolor{gray!10}{-0.804} & \cellcolor{gray!10}{-90.042} & \cellcolor{gray!10}{1} & \cellcolor{gray!10}{0.36}\\
\em{S. chilense} & LA1782 & -15.267 & -74.633 & 1000 & 0.35\\
\em{\cellcolor{gray!10}{S. chilense}} & \cellcolor{gray!10}{LA4117A} & \cellcolor{gray!10}{-22.907} & \cellcolor{gray!10}{-67.941} & \cellcolor{gray!10}{3540} & \cellcolor{gray!10}{0.22}\\
\addlinespace
\em{S. chmielewskii} & LA1028 & -13.883 & -73.017 & 3000 & 0.43\\
\em{\cellcolor{gray!10}{S. chmielewskii}} & \cellcolor{gray!10}{LA1316} & \cellcolor{gray!10}{-13.400} & \cellcolor{gray!10}{-73.906} & \cellcolor{gray!10}{2920} & \cellcolor{gray!10}{1.07}\\
\em{S. corneliomulleri} & LA0107 & -13.117 & -76.383 & 60 & 0.02\\
\em{\cellcolor{gray!10}{S. corneliomulleri}} & \cellcolor{gray!10}{LA0444} & \cellcolor{gray!10}{-13.433} & \cellcolor{gray!10}{-76.133} & \cellcolor{gray!10}{100} & \cellcolor{gray!10}{0.49}\\
\em{S. galapagense} & LA0436 & -0.953 & -90.978 & 40 & 0.20\\
\addlinespace
\em{\cellcolor{gray!10}{S. galapagense}} & \cellcolor{gray!10}{LA1044} & \cellcolor{gray!10}{-0.284} & \cellcolor{gray!10}{-90.548} & \cellcolor{gray!10}{0} & \cellcolor{gray!10}{0.18}\\
\em{S. habrochaites} & LA0407 & -2.181 & -79.884 & 70 & 0.47\\
\em{\cellcolor{gray!10}{S. habrochaites}} & \cellcolor{gray!10}{LA1777} & \cellcolor{gray!10}{-9.550} & \cellcolor{gray!10}{-77.700} & \cellcolor{gray!10}{3216} & \cellcolor{gray!10}{0.42}\\
\em{S. huaylasense} & LA1358 & -9.533 & -77.967 & 750 & 0.58\\
\em{\cellcolor{gray!10}{S. huaylasense}} & \cellcolor{gray!10}{LA1360} & \cellcolor{gray!10}{-9.546} & \cellcolor{gray!10}{-77.929} & \cellcolor{gray!10}{1490} & \cellcolor{gray!10}{0.48}\\
\addlinespace
\em{S. huaylasense} & LA1364 & -10.133 & -77.383 & 2920 & 0.88\\
\em{\cellcolor{gray!10}{S. lycopersicoides}} & \cellcolor{gray!10}{LA2951} & \cellcolor{gray!10}{-19.317} & \cellcolor{gray!10}{-69.450} & \cellcolor{gray!10}{2200} & \cellcolor{gray!10}{0.49}\\
\em{S. lycopersicoides} & LA4126 & -19.287 & -69.396 & 3120 & 0.39\\
\em{\cellcolor{gray!10}{S. neorickii}} & \cellcolor{gray!10}{LA1322} & \cellcolor{gray!10}{-13.483} & \cellcolor{gray!10}{-72.442} & \cellcolor{gray!10}{2380} & \cellcolor{gray!10}{0.68}\\
\em{S. neorickii} & LA2133 & -3.400 & -79.183 & 1980 & 1.05\\
\addlinespace
\em{\cellcolor{gray!10}{S. pennellii}} & \cellcolor{gray!10}{LA0716} & \cellcolor{gray!10}{-16.225} & \cellcolor{gray!10}{-73.617} & \cellcolor{gray!10}{50} & \cellcolor{gray!10}{0.19}\\
\em{S. pennellii} & LA0750 & -14.775 & -75.034 & 550 & 0.05\\
\em{\cellcolor{gray!10}{S. pennellii}} & \cellcolor{gray!10}{LA3778} & \cellcolor{gray!10}{-14.775} & \cellcolor{gray!10}{-75.034} & \cellcolor{gray!10}{616} & \cellcolor{gray!10}{0.05}\\
\em{S. peruvianum} & LA2744 & -18.550 & -70.150 & 400 & 0.16\\
\em{\cellcolor{gray!10}{S. peruvianum}} & \cellcolor{gray!10}{LA2964} & \cellcolor{gray!10}{-18.028} & \cellcolor{gray!10}{-70.835} & \cellcolor{gray!10}{75} & \cellcolor{gray!10}{1.16}\\
\addlinespace
\em{S. pimpinellifolium} & LA1269 & -11.483 & -77.075 & 400 & 0.50\\
\em{\cellcolor{gray!10}{S. pimpinellifolium}} & \cellcolor{gray!10}{LA1589} & \cellcolor{gray!10}{-8.433} & \cellcolor{gray!10}{-78.817} & \cellcolor{gray!10}{30} & \cellcolor{gray!10}{0.11}\\
\em{S. pimpinellifolium} & LA2933 & -1.442 & -80.562 & 375 & 1.62\\
\em{\cellcolor{gray!10}{S. sitiens}} & \cellcolor{gray!10}{LA4116} & \cellcolor{gray!10}{-22.159} & \cellcolor{gray!10}{-68.782} & \cellcolor{gray!10}{2960} & \cellcolor{gray!10}{0.06}\\
\bottomrule

@@

\subsubsection{\texorpdfstring{Cleaning \aqcurve{}
curves}{Cleaning  curves}}\label{cleaning-curves-1}

The raw data set consisted of 658 \aqcurve{} curves with an average of
19.4. Manual curation of a data set this size in a principled,
consistent manner is not feasible. Therefore, we automated data cleaning
using custom \emph{R} scripts. Cleaning is divided into two sequential
steps (\autoref{tbl-cleaning2}).

\caption{\label{tbl-cleaning2}Two sequential steps for cleaning
\agcurve{} curves. The rationale and procedure for each step are
described in the text. The rightmost columns summarize the number of
curves and mean number of points per curve remaining after each step.}

\centering
\begin{tabular}{>{\raggedright\arraybackslash}p{9cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
\toprule
Step: description & Number of curves & Number of points per curve\\
\midrule
\cellcolor{gray!10}{1. remove outliers within each \aqcurve{} curve} & \cellcolor{gray!10}{658} & \cellcolor{gray!10}{19.1}\\
2. remove \aqcurve{} curves with poor fit & 652 & 19.1\\
\bottomrule
\end{tabular}

\paragraph{\texorpdfstring{Remove outliers within each \aqcurve{}
curve}{Remove outliers within each  curve}}\label{remove-outliers-within-each-curve-1}

\emph{Procedure}: We fit provisional nonrectangular hyperbola
(\emph{59}) to each \aqcurve{} curve using nonlinear regression with the
\texttt{nlsLM()} function from the \emph{R} package \textbf{minpack.lm}
version 1.2.4 (\emph{60}). We sequentially removed data points with an
absolute external studentized residual \(> 3\) until none remained.

\paragraph{\texorpdfstring{Remove \aqcurve{} curves with poor
fit}{Remove  curves with poor fit}}\label{remove-curves-with-poor-fit}

\emph{Rationale}: \aqcurve{} curves with a poor fit to the
nonrectangular hyperbola most likely indicate systematic measurement
error and/or the leaf was not fully acclimated to the chamber
environment.

\emph{Procedure}: As described above, we fit provisional nonrectangular
hyperbola to each \aqcurve{} curve and calculated the model \(r^2\).
There was a clear break between typical curves and poorly fitting curves
where \(r^2 < 0.99\). We therefore removed \aqcurve{} curves with
\(r^2 < 0.99\).

\subsubsection{\texorpdfstring{Bayesian data analysis in
\emph{Stan}}{Bayesian data analysis in Stan}}\label{bayesian-data-analysis-in-stan}

We fit five models to test predictions of competing hypotheses about why
amphistomy advantage (\aax) might be greater for leaves in sunny, open
habitats. This section provides an overview of differences among models
(\autoref{tbl-models}). The next sections describe how we fit models in
\emph{Stan}, all model parameters and priors, and specific predictions
about parameter values for each hypothesis.

\caption{\label{tbl-models}Summary of differences among competing models
of how \aax{} varies with light intensity, light treatment, and among
populations as a function of native \ppfd. The models are numbered from
simpler to more complex. All models include fixed effects of light
intensity and light treatment; some models include interactions between.
All models include a phylogenetic random effect of population on \aax;
some models include varying effects of light intensity and light
treatment among populations. The last column indicates the
population-level \aax{} variable we used as a response to native \ppfd.}

\begin{tabular}{>{\centering\arraybackslash}p{0.3in}>{\raggedright\arraybackslash}p{1.3in}>{\raggedright\arraybackslash}p{3.4in}>{\raggedright\arraybackslash}p{1in}}
\toprule
Model & Fixed effects & Phylogenetic random effects & Response to native \ppfd\\
\midrule
1 & light intensity & varying intercept among populations & $\mathrm{AA}_{0, \text{acc}}$\\

 & light treatment &  & \\
\cmidrule{1-4}
2 & light intensity & varying intercept among populations & $\mathrm{AA}_{0, \text{acc}}$\\

 & light treatment &  & \\

 & intensity $\times$ treatment &  & \\
\cmidrule{1-4}
3 & light intensity & varying intercept among populations & $\mathrm{AA}_{2000, \text{acc}}$\\

 & light treatment & varying effect of high light intensity among populations & \\
\cmidrule{1-4}
4 & light intensity & varying intercept among populations & $\mathrm{AA}_{\text{sun, acc}}$\\

 & light treatment & varying effect of sun treatment among populations & \\
\cmidrule{1-4}
5 & light intensity & varying intercept among populations & $\mathrm{AA}_{2000, \text{sun, acc}}$\\

 & light treatment & varying effect of high light intensity among populations & \\

 & intensity $\times$ treatment & varying effect of sun treatment among populations & \\
\bottomrule
\end{tabular}

\paragraph{\texorpdfstring{Fitting models in
\emph{Stan}}{Fitting models in Stan}}\label{sec-fitting}

We fit Bayesian models with MCMC sampling in the probabilistic
programming language \emph{Stan} (\emph{61}) using the \emph{R} package
\textbf{brms} version 2.22.0 (\emph{62}). We used \emph{CmdStan} version
2.35.0 and \textbf{cmdstanr} version 0.9.0 (\emph{63}) to interface with
\emph{R} version 4.5.0 (\emph{64}). We sampled the posterior
distribution from 1 chains with 2000 iterations each after 2000 warmup
iterations per chain. We estimated parameters and confidence intervals
as the median and 95\% quantile intervals of the posterior,
respectively. We chose the number of chains, warmup and sampling
iterations, and maximum treedepth so that parameter estimates converged
(\(\hat{R} < 1.01\) (\emph{65})) and the effective sample size (ESS) for
each parameter was \(> 400\).

\paragraph{Parameter estimation and priors}\label{sec-parameters}

There were four levels of parameter estimation in our analysis:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Estimate \agcurve{} curve parameters
\item
  Estimate \aax{} for each light intensity with leaf using \agcurve{}
  curve parameters
\item
  Estimate the effects of light intensity, light treatment, and
  population on \aax{} (assimilatory and plasticity hypotheses)
\item
  Estimate the effects of native light habitat on population-level
  \aax{} (constitutive hypothesis)
\end{enumerate}

Although the higher-level parameter estimates depend on the lower-level
parameter estimates, we fit all models simultaneously to ensure that the
uncertainty in lower-level estimates propagated to higher-level
estimates.

\begin{longtable}{>{\raggedright\arraybackslash}p{1in}>{\raggedright\arraybackslash}p{5in}}

\caption{\label{tbl-parameters}Description of parameters estimated in
the hierarchical Bayesian model. The \texttt{Parameter} column lists the
parameter name as it appears in text. The \texttt{Description} column
provides a brief description of the parameter.}

\toprule
Parameter & Description\\
\midrule
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{\agcurve{} curve parameters}}\\
\hspace{1em}\cellcolor{gray!10}{$\mathbf{B}_\text{curve}$} & \cellcolor{gray!10}{$n_\text{curve} \times 3$ array of random \agcurve{} curve-level coefficients ($b_{0,j}$, $b_{1,j}$, $b_{2,j}$); $\mathbf{B}_\text{curve} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_\text{curve})$}\\
\hspace{1em}$\vec{\beta}_\text{curve}$ & vector of mean quadratic coefficients ($\beta_0$, $\beta_1$, $\beta_2$)\\
\hspace{1em}\cellcolor{gray!10}{$\symbf{\Sigma}_\text{curve}$} & \cellcolor{gray!10}{$3 \times 3$ covariance matrix of curve-level coefficients}\\
\hspace{1em}$\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}$ & minimum residual standard deviation when the measured leaf surface area is $\qty{6}{\raiseto{2}\centi\meter}$\\
\hspace{1em}\cellcolor{gray!10}{$\beta_{S,\epsilon}$} & \cellcolor{gray!10}{slope of the relationship between residual standard deviation and measured leaf surface area (log-link scale)}\\
\hspace{1em}$\rho_{\epsilon}$ & lag-1 residual autocorrelation\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{\aax{} for each light intensity with leaf using agcurve{} curve parameters}}\\
\hspace{1em}\cellcolor{gray!10}{$\widehat{\mathrm{AA}}_{klm}$} & \cellcolor{gray!10}{estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in accession $m$}\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{effects of light intensity, light treatments, and population on \aax}}\\
\hspace{1em}$\beta_{\mathrm{AA},0}$ & intercept of \aax{} at low light intensity in shade treatment\\
\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},2000}$} & \cellcolor{gray!10}{effect of high light intensity at \ppfdequals{2000} on \aax}\\
\hspace{1em}$\beta_{\mathrm{AA},\text{sun}}$ & effect of sun treatment on \aax{}\\
\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},2000,\text{sun}}$} & \cellcolor{gray!10}{effect of high light intensity at \ppfdequals{2000} on \aax{} in sun treatment}\\
\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{acc}}$ & vector of $n_\text{acc}$ phylogenetically structured random accession-level effects on \aax; $\vec{\beta}_{\mathrm{AA},\text{acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\text{AA,acc}})$\\
\hspace{1em}\cellcolor{gray!10}{$\symbf{\Sigma}_{\text{AA,acc}}$} & \cellcolor{gray!10}{$n_\text{acc} \times n_\text{acc}$ covariance matrix of phylogenetically structured random accession-level effects on \aax;}\\
\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{rep}}$ & vector of $n_\text{rep}$ random replicate-level effects on \aax; $\vec{\beta}_{\mathrm{AA},\text{rep}} \sim \text{Normal}(0, \sigma_{\mathrm{AA},\text{rep}})$\\
\hspace{1em}\cellcolor{gray!10}{$\vec{\beta}_{\mathrm{AA},\text{2000,acc}}$} & \cellcolor{gray!10}{vector of $n_\text{acc}$ phylogenetically structured random accession-level effects of high light intensity at \ppfdequals{2000} on \aax; $\vec{\beta}_{\mathrm{AA},\text{2000,acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\mathrm{AA},\text{2000,acc}})$}\\
\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{sun,acc}}$ & vector of $n_\text{acc}$ phylogenetically structured random accession-level effects of sun treatment on \aax; $\vec{\beta}_{\mathrm{AA},\text{sun,acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\mathrm{AA},\text{sun,acc}})$\\

>>>>

We tested these hypotheses by comparing \aax{} among amphistomatous wild
tomato species (\emph{51}) from different native light habitats, grown
under simulated sun and shade light treatments, and measured under
contrasting light intensity (). We measured AA on 572 individual plants
from 29 populations (average of 9.86 replicates per light treatment)
using a recently developed method (\emph{46}). With this method, we
directly compare the photosynthetic rate of an untreated amphistomatous
leaf to that of the same leaf with gas exchange blocked through the
adaxial (upper) surface by transparent plastic, which we refer to as
`pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at
identical whole-leaf \(g_\text{sc}\), we measure \(A\) over a range of
\(g_\text{sc}\), inducing stomatal opening and closure by modulating
humidity (see Materials and Methods for further details). We estimated
`amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but with
modifications previously described in (\emph{46}) and here (Materials
and Methods). The native light intensity was represented by plant area
index (PAI \(\unit{\meter\squared\per\meter\squared}\)), estimated using
a global gridded data set derived from the Global Ecosystem Dynamics
Investigation {[}GEDI; (\emph{52}){]} and georeferenced accession
collection information from the Tomato Genetics Resource Center. The
growth light intensities were \ppfdequals{761} (sun treatment) and
\ppfdqty{115} (shade treatment) while all other environment conditions
were nearly identical (see materials and methods in supplementary
materials). The high and low measurements intensities were
\ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150} (87.0:13.0
red:blue), respectively.

Consistent with biophysical theory of CO\(_2\) diffusion within leaves,
\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). Bayesian
phylogenetic mixed effects models that allowed \aax{} to vary between
measurement light intensities, growth light intensities, and among
populations outperformed simpler models based on their information
criteria (Table~\ref{tbl-aa_loo1}). Measured under high light intensity,
\aax{} was consistently greater for sun plants. The average \aax{} among
populations in the shade treatment was 0.041 (range: 0.010--0.113; 21 of
29 populations significant); however, the same populations grown at high
light intensity showed a mean \aax{} of 0.053 (range: 0.021--0.125; 19
of 29 populations significant). Contrary to the predictions of the
assimilatory hypothesis, \aax{} was greater in all populations under low
measurement light intensity for both sun and shade grown plants. The
effect of low light on \aax{} was more pronounced in the sun-grown
plants, where \aax{} was significantly greater under low measurement
light intensity in 11 of 29 populations compared to to 4 populations for
shade-grown plants. The overall average \aax{} of shade and sun grown
plants measured under low light intensity was 0.064 (range:
0.017--0.132; 28 of 29 populations significant) and 0.094 (range:
0.036--0.197; 24 of 29 populations significant), respectively. There was
a slight tendency for populations from more closed habitats (greater
PAI), but \aax{} varied widely among populations from open habitats (low
PAI) regardless of growth and measurement light intensities
(Fig.~\ref{fig-aa}B).

@@

maybe this wording is better: Amphistomy likely plays a significant, but
under appreciated, role in the carbon and water balance of plant groups
where this trait is common, such as herbaceous plants in sunny habitats
and woody species in arid biomes (cites). If our estimates of AA from
the wild tomato clades

note: estimates of AA are less than that described by Parkhurst and Mott
(1990) and also less than that of Xiong and Flexas (2021). What does
this tell us?

@@

\toprule
Species & TGRC accession & Latitude & Longitude & Elevation (mas) & $\mathrm{PAI}~(\unit{\meter\squared\per\meter\squared})$\\
\midrule
\endfirsthead
\multicolumn{6}{@{}l}{\textit{(continued)}}\\
\toprule
Species & TGRC accession & Latitude & Longitude & Elevation (mas) & $\mathrm{PAI}~(\unit{\meter\squared\per\meter\squared})$\\
\midrule
\endhead

\endfoot
\bottomrule
\endlastfoot
\em{\cellcolor{gray!10}{S. arcanum}} & \cellcolor{gray!10}{LA2172} & \cellcolor{gray!10}{-6.008} & \cellcolor{gray!10}{-78.858} & \cellcolor{gray!10}{662} & \cellcolor{gray!10}{0.70}\\
\em{S. cheesmaniae} & LA0429 & -0.644 & -90.329 & 800 & 0.60\\
\em{\cellcolor{gray!10}{S. cheesmaniae}} & \cellcolor{gray!10}{LA3124} & \cellcolor{gray!10}{-0.804} & \cellcolor{gray!10}{-90.042} & \cellcolor{gray!10}{1} & \cellcolor{gray!10}{0.36}\\
\em{S. chilense} & LA1782 & -15.267 & -74.633 & 1000 & 0.35\\
\em{\cellcolor{gray!10}{S. chilense}} & \cellcolor{gray!10}{LA4117A} & \cellcolor{gray!10}{-22.907} & \cellcolor{gray!10}{-67.941} & \cellcolor{gray!10}{3540} & \cellcolor{gray!10}{0.22}\\
\addlinespace
\em{S. chmielewskii} & LA1028 & -13.883 & -73.017 & 3000 & 0.43\\
\em{\cellcolor{gray!10}{S. chmielewskii}} & \cellcolor{gray!10}{LA1316} & \cellcolor{gray!10}{-13.400} & \cellcolor{gray!10}{-73.906} & \cellcolor{gray!10}{2920} & \cellcolor{gray!10}{1.07}\\
\em{S. corneliomulleri} & LA0107 & -13.117 & -76.383 & 60 & 0.02\\
\em{\cellcolor{gray!10}{S. corneliomulleri}} & \cellcolor{gray!10}{LA0444} & \cellcolor{gray!10}{-13.433} & \cellcolor{gray!10}{-76.133} & \cellcolor{gray!10}{100} & \cellcolor{gray!10}{0.49}\\
\em{S. galapagense} & LA0436 & -0.953 & -90.978 & 40 & 0.20\\
\addlinespace
\em{\cellcolor{gray!10}{S. galapagense}} & \cellcolor{gray!10}{LA1044} & \cellcolor{gray!10}{-0.284} & \cellcolor{gray!10}{-90.548} & \cellcolor{gray!10}{0} & \cellcolor{gray!10}{0.18}\\
\em{S. habrochaites} & LA0407 & -2.181 & -79.884 & 70 & 0.47\\
\em{\cellcolor{gray!10}{S. habrochaites}} & \cellcolor{gray!10}{LA1777} & \cellcolor{gray!10}{-9.550} & \cellcolor{gray!10}{-77.700} & \cellcolor{gray!10}{3216} & \cellcolor{gray!10}{0.42}\\
\em{S. huaylasense} & LA1358 & -9.533 & -77.967 & 750 & 0.58\\
\em{\cellcolor{gray!10}{S. huaylasense}} & \cellcolor{gray!10}{LA1360} & \cellcolor{gray!10}{-9.546} & \cellcolor{gray!10}{-77.929} & \cellcolor{gray!10}{1490} & \cellcolor{gray!10}{0.48}\\
\addlinespace
\em{S. huaylasense} & LA1364 & -10.133 & -77.383 & 2920 & 0.88\\
\em{\cellcolor{gray!10}{S. lycopersicoides}} & \cellcolor{gray!10}{LA2951} & \cellcolor{gray!10}{-19.317} & \cellcolor{gray!10}{-69.450} & \cellcolor{gray!10}{2200} & \cellcolor{gray!10}{0.49}\\
\em{S. lycopersicoides} & LA4126 & -19.287 & -69.396 & 3120 & 0.39\\
\em{\cellcolor{gray!10}{S. neorickii}} & \cellcolor{gray!10}{LA1322} & \cellcolor{gray!10}{-13.483} & \cellcolor{gray!10}{-72.442} & \cellcolor{gray!10}{2380} & \cellcolor{gray!10}{0.68}\\
\em{S. neorickii} & LA2133 & -3.400 & -79.183 & 1980 & 1.05\\
\addlinespace
\em{\cellcolor{gray!10}{S. pennellii}} & \cellcolor{gray!10}{LA0716} & \cellcolor{gray!10}{-16.225} & \cellcolor{gray!10}{-73.617} & \cellcolor{gray!10}{50} & \cellcolor{gray!10}{0.19}\\
\em{S. pennellii} & LA0750 & -14.775 & -75.034 & 550 & 0.05\\
\em{\cellcolor{gray!10}{S. pennellii}} & \cellcolor{gray!10}{LA3778} & \cellcolor{gray!10}{-14.775} & \cellcolor{gray!10}{-75.034} & \cellcolor{gray!10}{616} & \cellcolor{gray!10}{0.05}\\
\em{S. peruvianum} & LA2744 & -18.550 & -70.150 & 400 & 0.16\\
\em{\cellcolor{gray!10}{S. peruvianum}} & \cellcolor{gray!10}{LA2964} & \cellcolor{gray!10}{-18.028} & \cellcolor{gray!10}{-70.835} & \cellcolor{gray!10}{75} & \cellcolor{gray!10}{1.16}\\
\addlinespace
\em{S. pimpinellifolium} & LA1269 & -11.483 & -77.075 & 400 & 0.50\\
\em{\cellcolor{gray!10}{S. pimpinellifolium}} & \cellcolor{gray!10}{LA1589} & \cellcolor{gray!10}{-8.433} & \cellcolor{gray!10}{-78.817} & \cellcolor{gray!10}{30} & \cellcolor{gray!10}{0.11}\\
\em{S. pimpinellifolium} & LA2933 & -1.442 & -80.562 & 375 & 1.62\\
\em{\cellcolor{gray!10}{S. sitiens}} & \cellcolor{gray!10}{LA4116} & \cellcolor{gray!10}{-22.159} & \cellcolor{gray!10}{-68.782} & \cellcolor{gray!10}{2960} & \cellcolor{gray!10}{0.06}\\*

@@

\subsubsection{\texorpdfstring{Bayesian data analysis with
\emph{Stan}}{Bayesian data analysis with Stan}}\label{sec-analysis}

Our analysis pipeline consisted of four main steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Estimate \agcurve{} curve parameters for each leaf
\item
  Estimate \aax{} for both measurement light intensities using
  \agcurve{} curve parameters
\item
  Estimate the effects of light intensity, light treatment, and
  population on \aax{} (assimilatory and plasticity hypotheses)
\item
  Estimate the effects of native light habitat on population-level
  \aax{} (constitutive hypothesis)
\end{enumerate}

All models were using a Bayesian model with HMC sampling in the
probabilistic programming language \emph{Stan} (\emph{59}) using the
\emph{R} package \textbf{brms} version 2.22.0 (\emph{60}). We used
\emph{CmdStan} version 2.36.0 and \textbf{cmdstanr} version 0.9.0
(\emph{61}) to interface with \emph{R} version 4.5.0 (\emph{62}). We
sampled the posterior distribution from a single chain with a minimum of
1000 iterations after 1000 warmup iterations. If necessary, we refit
models with more iterations to decrease the Gelman-Rubin convergence
statistic (\(\hat{R}\)) (\emph{63}) to less than
\(`aa_stats[["max_rhat"]]`\) and the bulk effective sample size (ESS) to
greater than \(`aa_stats[["min_ess"]]`\) for all model parameters. We
also increased the the \texttt{adapt\_delta} parameter until there were
fewer than \(`aa_stats[["n_divergent"]]`\) divergent transitions.

\paragraph{\texorpdfstring{\agcurve{} curve
parameters}{ curve parameters}}\label{curve-parameters}

We modeled \logA as a quadratic function of \loggsw for each leaf,
measured at low and high light intensity, in both amphistomatous
(untreated) and pseudohypostomatous (propafilm treatment) states.

\paragraph{\texorpdfstring{\aax{} for each light intensity within leaf
using \agcurve{} curve
parameters}{ for each light intensity within leaf using  curve parameters}}\label{for-each-light-intensity-within-leaf-using-curve-parameters}

\subparagraph{MARKER - add figure of
this}\label{marker---add-figure-of-this}

We estimated \aax{} for each light intensity by integrating the
difference in \logA{} between the amphi and pseudohypo \agcurve{} curves
over the range of \gsw{} values where the curves overlap (from
\(\min(\log(g_\text{sw}))\) to \(\max(\log(g_\text{sw}))\)).

\[\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx\]
where \(\theta_\text{amphi}\) and \(\theta_\text{hypo}\) are the
quadratic parameters of the amphi and pseudohypo \agcurve{} curves,
respectively. We calculated \(\widehat{\mathrm{AA}}\) from each draw of
the posterior for the paired amphi and pseudohypo \agcurve{} curves to
obtain the posterior distribution of \aax{} for each leaf at each light
intensity. We then summarized the posterior distribution of \aax{} for
each leaf at each light intensity by the posterior median, which became
our estimate of \aax{} for subsequent analyses.

\subsubsection{Bayesian phylogenetic mixed effects
models}\label{bayesian-phylogenetic-mixed-effects-models}

We fit Bayesian mixed effects models with phylogenetically structured
random effects to test predictions of competing hypotheses about why
amphistomy advantage (\aax) might be greater for leaves in sunny, open
habitats. We fit all models combinations of measurement light intensity,
growth light intensity, and their interaction as both fixed effects and
random effects among populations. In other words, we test whether there
were main effects of light intensity treatments and/or whether
populations varied in their response to light intensity treatments. In
all models, we included population as a phylogenetically structured
random effect. We crossed those fixed and random effect structures with
fixed effects of measurement light intensity and/or growth light
intensity on the distributional parameter, \(\sigma\), the residual
variance in \aax{}. In total, we fit 100 models, each with a different
combination of fixed, random, and distributional effects. We used the
RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb
genomic windows from (\textbf{pease\_phylogenomics\_2016?}). Two of our
populations were not in this tree. We used accession LA1044 in place of
LA3909, two populations of \emph{S. galapagnse}; LA0750 was added as
sister to LA0716, two closely replated populations of \emph{S.
pennellii}. The node was placed half-way between the next deepest node.

NEED TO EDIT THIS: The next sections describe how we fit models in
\emph{Stan}, all model parameters and priors, and specific predictions
about parameter values for each hypothesis. Each curve was fit For
initial fits, we sampled the posterior distribution from a single chain
with 1000 iterations after 1000 warmup iterations. After model selection
(see below), we refit the selected model with more iterations if
necessary to decrease the Gelman-Rubin convergence statistic
(\(\hat{R}\)) (\emph{63}) to less than \(`aa_stats[["max_rhat"]]`\) and
the bulk effective sample size (ESS) to greater than
\(`aa_stats[["min_ess"]]`\) for all model parameters.

\paragraph{\texorpdfstring{Fitting models in \emph{Stan} using
\textbf{brms}
interface}{Fitting models in Stan using brms interface}}\label{sec-fitting}

We fit Bayesian models with HMC sampling in the probabilistic
programming language \emph{Stan} (\emph{59}) using the \emph{R} package
\textbf{brms} version 2.22.0 (\emph{60}). We used \emph{CmdStan} version
2.36.0 and \textbf{cmdstanr} version 0.9.0 (\emph{61}) to interface with
\emph{R} version 4.5.0 (\emph{62}). For initial fits, we sampled the
posterior distribution from a single chain with 1000 iterations after
1000 warmup iterations. After model selection (see below), we refit the
selected model with more iterations if necessary to decrease the
Gelman-Rubin convergence statistic (\(\hat{R}\)) (\emph{63}) to less
than \(`aa_stats[["max_rhat"]]`\) and the bulk effective sample size
(ESS) to greater than \(`aa_stats[["min_ess"]]`\) for all model
parameters.
